\pdftitle

### 2017-05-15 maxNt.tex
more discussion

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-15T14:30:55+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage{datetime2}
\DTMnewdatestyle{mydate}%
{% definitions
\renewcommand*{\DTMdisplaydate}[4]{%
\number##3\ \DTMenglishmonthname{##2} ##1}%
\renewcommand*{\DTMDisplaydate}{\DTMdisplaydate}%
}
\DTMsetdatestyle{mydate}

@@

Suppose our desired constraints are $\expe{\yxs } = c_1$ and
$\expe{\yxxs} = c_2$, the expectations being given by the fourth term of
\eqn~\eqref{eq:expe_products}. Applying \me\ to the distribution of the
average of the full population we find
\begin{equation}
  \label{eq:full_pop_av_example_ME}
  \p(\yXf = \yRf \| \yHa)  = \yL\,
\binom{N}{N \yRf}\exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr]
\end{equation}
with such parameters $\yL$, $\yL_1$, $\yL_2$ as to satisfy normalization
and constraints. We call this state of knowledge $\yHa$. Marginalization by
\eqn~\eqref{eq:marginal} then gives
\begin{equation}
  \label{eq:full_pop_av_example_ME_marg}
  \p(\yxs = \yrs \| \yHa)  = \yL
  \sum_{N\yRf=0}^N
  \binom{n}{n\yrs} \binom{N-n}{N\yRf-n\yrs} \exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr].
\end{equation}
Applying \me\ at the sample level, using the third term of
\eqn~\eqref{eq:expe_products}, we find
\begin{equation}
  \label{eq:sample_av_example_ME}
  \p(\yxs = \yrs \| \yHb)  = \yl\,
  \binom{n}{n\yrs}
  \exp\biggl[\yl_2\, \frac{\yrs\,(n \yrs-1)}{n-1} + \yl_1\, \yrs\biggr]
\end{equation}
with such parameters $\yl$, $\yl_1$, $\yl_2$ as to satisfy the constraints.
We call this state of knowledge $\yHb$. For example, with $N=5\,000$,
$n=100$, $c_1=0.45$, $c_2=0.35$, we find $\yL_1 = -13\,321.9$,
$\yL_2 = 13\,321.5$, $\yl_1=-269.3$, $\yl_2 = 268.9$, and appropriate
normalization constants. The resulting distributions
$\p(\yxs = \yrs \| \yHa)$, $\p(\yxs = \yrs \| \yHb)$ are shown in
\fig~\ref{fig:diff_maxent_pop_sample}; they are clearly different.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_100.pdf}%
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.45$,
  $\expe{\yxxs \|\yH}=0.35$, and $N=5\,000$, $n=100$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb

@@

The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases of interest, especially with numerically
small constraints. One could thus be used as an approximation of the other
depending on the circumstances -- \eg\ if probability tails are
unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5\,000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb

\mynote{We don't give an answer. Different purposes. Bright side: no real
  contradiction because we are expressing two states of knowledge, not
  physical situations. Also, in some case numerically very similar results.

  The distribution obtained from the population-level application has lower
  entropy; this usually means that in includes additional assumptions than
  the other. Intuitively, this additional assumption is that the $n$
  neurons are part of a larger population, of which they are
  representative.

  Same phenomenon in statistical mechanics \cite{maesetal1999}. Somehow
  similar quandary for equilibrium state at the end of non-equilibrium
  process: Gibbs state vs initial Gibbs state + Liouville
  evolution \cite[\sect~4]{jaynes1985d_r1993}; but in this case both
  distributions are immensely sharp, and this doesn't affect predictions}

\bigskip
{\centering\mynote{*** From here down old text to be discarded ***}\par}
\section{***}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-15T15:53:25+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage{wrapfig}
\usepackage{datetime2}
\DTMnewdatestyle{mydate}%
{% definitions
\renewcommand*{\DTMdisplaydate}[4]{%
\number##3\ \DTMenglishmonthname{##2} ##1}%
\renewcommand*{\DTMDisplaydate}{\DTMdisplaydate}%
}
\DTMsetdatestyle{mydate}

@@

Suppose our desired constraints are $\expe{\yxs } = c_1$ and
$\expe{\yxxs} = c_2$, the expectations being given by the fourth term of
\eqn~\eqref{eq:expe_products}. Applying \me\ to the distribution of the
average of the full population we find
\begin{equation}
  \label{eq:full_pop_av_example_ME}
  \p(\yXf = \yRf \| \yHa)  = \yL\,
\binom{N}{N \yRf}\exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr]
\end{equation}
with such parameters $\yL$, $\yL_1$, $\yL_2$ as to satisfy normalization
and constraints. We call this state of knowledge $\yHa$. Marginalization by
\eqn~\eqref{eq:marginal} then gives
\begin{equation}
  \label{eq:full_pop_av_example_ME_marg}
  \p(\yxs = \yrs \| \yHa)  = \yL
  \sum_{N\yRf=0}^N
  \binom{n}{n\yrs} \binom{N-n}{N\yRf-n\yrs} \exp\biggl[
\yL_2\, \frac{\yRf\,(N \yRf-1)}{N-1} + \yL_1\, \yRf\biggr].
\end{equation}
Applying \me\ at the sample level, using the third term of
\eqn~\eqref{eq:expe_products}, we find
\begin{equation}
  \label{eq:sample_av_example_ME}
  \p(\yxs = \yrs \| \yHb)  = \yl\,
  \binom{n}{n\yrs}
  \exp\biggl[\yl_2\, \frac{\yrs\,(n \yrs-1)}{n-1} + \yl_1\, \yrs\biggr]
\end{equation}
%\setlength{\intextsep}{0.5ex}%
%\begin{wrapfigure}{r}{0.5\linewidth}
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\linewidth]{different_maxent_pop_sample_100.pdf}%
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.45$,
  $\expe{\yxxs \|\yH}=0.35$, and $N=5\,000$, $n=100$.}
\label{fig:diff_maxent_pop_sample}
\end{figure}%maxent_pop_or_sample.nb
%\end{wrapfigure}
with such parameters $\yl$, $\yl_1$, $\yl_2$ as to satisfy the constraints.
We call this state of knowledge $\yHb$. For example, with $N=5\,000$,
$n=100$, $c_1=0.45$, $c_2=0.35$, we find $\yL_1 = -13\,321.9$,
$\yL_2 = 13\,321.5$, $\yl_1=-269.3$, $\yl_2 = 268.9$, and appropriate
normalization constants. The resulting distributions
$\p(\yxs = \yrs \| \yHa)$, $\p(\yxs = \yrs \| \yHb)$ are shown in
\fig~\ref{fig:diff_maxent_pop_sample}; they are clearly different.

@@

The probability distribution for the sample average obtained by applying
\me\ at the population level has by construction a lower entropy than the
one obtained by application at the sample level. It says that some sample
states $\yx$ should be considered less likely than in the other
distribution. The lower entropy signals the presence of an additional
physical assumption behind the former distribution. Such assumption,
intuitively, is that the sample is part of a larger population, of which it
is representative. And this assumption may affect our predictions.

The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5\,000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that the possibility of using two different distributions is
not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A quandary somehow similar
to ours also appears in the statistical description of an equilibrium state
at the end of a non-equilibrium process: we can describe our knowledge
about it either by a Gibbs distribution, or by the Liouville-evolved Gibbs
distribution associated with equilibrium state at the beginning of the
process. The two descriptions differ -- even though the final physical
state is exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The appearance
of this difference is natural: in one case we can make sharper predictions
about the state thanks to our knowledge of its preceding dynamics.
\mynote{but in this case both distributions are immensely sharp, and this
  doesn't affect predictions}

\mynote{Importance of the formulae relating sample and population.}

\newrefsection
\bigskip
{\centering\mynote{*** From here down old text to be discarded ***}\par}
\section{***}
